---
title: "PA1_template.Rmd"
author: "Sergey Sambor"
date: "April 26, 2017"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries needed for this analysis
```{r message=FALSE}
library(ggplot2)
library(dplyr)
```

__tickMarks__ a simple assistant function for plotting
```{r}
# tickMarks returns n tick marks used for time axis
tickMarks <- function (intervals, n) {
    intervals[seq(1, length(intervals), length(intervals)/n )]
}
```

## Data
The variables included in this dataset are:  

* __steps__: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* __date__: The date on which the measurement was taken in YYYY-MM-DD format
* __interval__: Identifier for the 5-minute interval in which measurement was taken  

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
```{r}
unzip("../activity.zip", exdir = "../data")
data0 <- read.csv("../data/activity.csv", colClasses = c("numeric", "Date", "numeric"))
data0 <- mutate(data0, 
    interval = gsub('^(.{2})(.*)$', #split to 2 groups with 2 numbers each
                    '\\1:\\2', #insert a colon symbol
                    sprintf("%04d", interval ) )) #add leading zeros
data0$interval <- as.factor(data0$interval)
```
Now 

* __steps__: numeric (integer) value
* __date__: Date value
* __interval__: a factor, in a form of "01:00", "08:00", "14:00"  

## The total number of steps taken each day
Allmost always the number of steps each day lays in between 7500 and 17000 steps.
The most frequent total numbers are inside [9000;13000]
```{r warning=FALSE}
daysTotalSteps <- data0 %>% 
    group_by(date) %>% 
    summarize(totalSteps = sum(steps))
qplot(totalSteps, data = daysTotalSteps, binwidth=2000)
```

Whilst mean and median number are, respectively
```{r}
mean(daysTotalSteps$totalSteps, na.rm = TRUE)
median(daysTotalSteps$totalSteps, na.rm = TRUE)
```

## Average number of steps taken across the day
The most steps are taken between 8 and 9 hours.
```{r}
periodStepsAvg <- data0 %>% 
    group_by(interval) %>% 
    summarise(stepsAvg = mean(steps, na.rm = TRUE))
ggplot(periodStepsAvg, aes(x=interval, y=stepsAvg, group=1)) +
    geom_line() +
    scale_x_discrete(breaks = tickMarks(periodStepsAvg$interval, 12))
```

And the period with the highest average number is 8:35
```{r}
periodStepsAvg[which.max(periodStepsAvg$stepsAvg),]
```

## Imputing missing data
The total number of missing data is
```{r}
sum(is.na(data0$steps))
```

For imputing missing data we use the average value for the same 5-min interval during two months period.
```{r}
data1 <- mutate(data0, 
    steps = round(ifelse(#replace NAs with steps' average for this inverval
        is.na(steps),
        periodStepsAvg$stepsAvg[which(periodStepsAvg$interval %in% interval)], 
        steps
        )))
```

### The new histogram of the total number of steps taken each day has changed slightly
Now it's the [9000;11000] steps are clearly the most frequent total number per day
```{r}
daysTotalSteps <- data1 %>% 
    group_by(date) %>% 
    summarize(totalSteps = sum(steps))
qplot(totalSteps, data = daysTotalSteps, binwidth=2000)
```

Whilst mean and median number, respectively, almost the same
```{r}
periodStepsAvg <- data1 %>% 
    group_by(interval) %>% 
    summarise(stepsAvg = mean(steps, na.rm = TRUE))
mean(daysTotalSteps$totalSteps, na.rm = TRUE)
median(daysTotalSteps$totalSteps, na.rm = TRUE)
```

##  Activity patterns in weekdays and weekends
Weekends activity differs to weekdays activity. There's no such a big peak of activity between 8 and 9 hour, but there is more activity throughout the day. Also there's much more activity in the evening on the weekend comparing to weekdays.
```{r}
data1 <- mutate(data1, 
    daytype = as.factor(ifelse(
        grepl("Sun|Sat", weekdays(date)), 
        "weekend", 
        "weekday")))
weekends <- data1 %>% 
    group_by(daytype, interval) %>% 
    summarize(avg_steps = mean(steps))
ggplot(weekends, aes(x=interval, y=avg_steps, group=daytype))+
    facet_grid(daytype~.)+
    geom_line()+
    scale_x_discrete(breaks = tickMarks (weekends$interval, 24))
```